Characterizing the Role of the Human Virome in Colorectal Cancer

### 2017-03-14 doc/manuscript.Rmd
More major revisions.

Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world. Although the cause of colorectal cancer remains unclear, it has been strongly linked to colon bacterial communities. Viruses are another important component of the colon microbial community that have yet to be studied in colorectal cancer, despite their oncogenic potential. We evaluated the colon virome (virus community) role in colorectal cancer using a cohort of 90 human subjects with either healthy, precancerous, or cancerous colons. We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic methods to compare the virome's role to that of the bacterial community. We found that alpha and beta diversity metrics were insufficient for detecting an association between the virome and colorectal cancer, but virome-based classification models were highly associated with colorectal cancer. Bacteriophages, not eukaryotic viruses, made up the majority of the CRC-associated virome, suggesting the community was indirectly linked to colorectal cancer, modulating bacterial community structure and functionality. Phages with broader host ranges had a more significant role in cancer development. These results provide foundational evidence that phage communities are associated with colorectal cancer, and that broadly tropic phages play more significant roles. Because of its importance, the virome should be considered in future colorectal cancer studies, as well as other cancer types.

**Word Count**: 229

# Significance Statement
Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide. It's progression and severity is linked to colon bacterial communities. Little is known about the cancer-associated colon virus communities and their influence on bacteria. We characterize this influence of colon virus community structure on colorectal cancer in humans. The link between the colon virome and colorectal cancer was established by using virome-based machine learning models to accurately classify patient cancer status. The cancer-virus link was driven primarily by bacteriophages (bacterial viruses). Wider phage host ranges were more strongly linked to colorectal cancer. The results suggest an indirect role for the virome impacting colorectal cancer by modulating their associated bacterial community.

**Word Count**: 119

@@

# Introduction
Cancer remains a devastating and persistent plague on humanity. Although cancer is still a primary cause of morbidity and mortality worldwide, we have made considerable therapeutic progress in recent decades. Perhaps one of our most impactful advances has not been in cancer treatment, but rather in detecting cancers at early stages, thereby improving treatment efficacy. This has been evident in a variety of prominent cancer types, with one of the most notable being colorectal cancer.

Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo]. The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans will have died from the disease [@Siegel:2014jo]. Although it remains a major health problem, the impact of colorectal cancer has been reduced by improved screening and prevention efforts [@Zauber:2015cr; @Siegel:2014jo].

Development of colorectal cancer is a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e. adenoma) in the large intestine [@Fearon:2011go]. If left untreated, the adenoma will develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death. Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr]. Survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz]. Screening methods are effective, but their invasiveness has created a lack of compliance, creating a need for accurate, non-invasive screening methods. One such method is screening associated colon microbial communities.

Although the cause of colorectal cancer remains unclear, it has been strongly associated with colon bacterial communities [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb]. This association has allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja]. While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and understanding underlying etiology, bacteria are only a subset of the colon microbiome. Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer.

Due to their mutagenic abilities and their propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Additionally, because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages are potentially indirectly linked to cancer. The colon virome (the virus community of the colon) has the potential to impact health and disease (e.g. cancer), and has been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. We aim to take this field of research further by beginning to assess the role of the virome in colorectal cancer.

Here we present a study of the colorectal cancer virome, highlighting the viruses most associated with the cancer state and the virome's utility for prognosis and diagnosis. We report that, like the association between the bacterial community and colorectal cancer was driven primarily by *Fusobacterium*, the association between the virome and colorectal cancer was driven by broadly infectious, bacteriophage hubs within the phage-bacteria ecosystem. By creating effective classification models using virus community signatures, we were able to accurately classify stool samples as cancerous, precancerous, or healthy. The implications of these findings are threefold. *First*, this supports a biological role for the virome in colorectal cancer development and suggests that more than bacteria are involved in the process. *Second*, we present a supplementary, or even alternative, virus-based approach for classification modeling of colorectal cancer using stool samples. *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer. We expect this study will provide opportunities for continued research into the role of the virome in human cancer development.

@@

## The Colorectal Cancer Virome Cohort
The study cohort consisted of 90 human subjects, 30 of which served as healthy controls, 30 of which had adenoma lesions consistent with a precancerous state, and 30 which had carcinoma lesions consistent with colorectal cancer **(Figure \ref{sampleproc})**. Half of the stool was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques. The 16S rRNA gene sequences were reported in a previous publication but re-analyzed here [@Zackular:2014fba]. The other half of the stool samples were purified for virus like particles (VLPs) and consequent genomic DNA extraction, followed by shotgun metagenomic sequencing. The VLP purification allowed us to observe the *active virome* because we only sequence those viruses that are encapsulated.

Virus DNA was purified by re-suspending the stool in buffer and removing contaminating cells (e.g. human, bacteria, etc) by filtration followed by contaminating cell lysis and degradation of the released genomic DNA **(Figure \ref{sampleproc})**. The resulting genomic DNA was used to prepare a shotgun metagenomic sequencing library that was sequenced on the Illumina HiSeq4000 platform. Each run was performed with a blank control to detect any contaminants from reagents. Only one of the nine viral controls detected DNA, which was of a minimal concentration, providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**. As was expected, these controls were sparsely sequenced and were mostly removed while sub-sampling to even depths **(Figure \ref{qualcontrol} B)**.

The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**. Because contigs only represent genome fragments, we further clustered related contigs into operational genomic units (OGUs) with the majority containing hundreds of related contigs **(Figure \ref{contigqc} - \ref{clustercontigqc})**. These operational units, which are conceptually similar to the operational taxonomic units (OTUs) used in 16S analysis, allow us to study bacterial and phage entities as highly related genomic entities.

## Diversity is Insufficient for Virome-Based Cancer Classification
Microbiome and disease associations are often described as being of an altered diversity (i.e. "dysbiosis"). We therefore initially evaluated the diversity of virome OGUs and their association with colorectal cancer. We utilized the Bray-Curtis dissimilarity metric to evaluate the differences in communities between disease states. To control for uneven sequencing depths, we subsampled to a minimum depth that maintained most samples while excluding the sparsely sequenced negative controls.

There was no statistically significant clustering of the disease groups, as visualized by NMDS ordination (Anosim p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu} A**). An Anosim test with a post hoc multivariate Tukey test was used to calculate the statistical significance of the differences between the disease groups based on the variance around the cluster centroids **(Figure \ref{betaogu} B**). There were no significant differences between the disease groups, although there was a strongly significant difference between the blank controls (those few that remained after sub-sampling) and the rest of the study groups, further supporting the quality of our sample set (Anosim p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**.

In addition to beta diversity, we also calculated the differences in virus alpha diversity associated with colorectal cancer. Again we found no significant alterations in either Shannon entropy or richness of the virus communities **(Figure \ref{betaogu} C-D)**. Overall, standard diversity metrics were insufficient for capturing the differences in the virus communities between disease states. This suggested to us that a more sophisticated approach for understanding the microbial community may be required, such as machine learning classification.

## Machine Learning Links the Virome to Colorectal Cancer
Previous work has shown that 16S rRNA community signatures are effective for classifying stool samples as originating from healthy, precancerous, or cancerous individuals [@Zackular:2014fba; @Baxter:2016dja]. This is valuable because it presents a potential alternative screening approach to an invasive and expensive colonoscopy. The exceptional performance of bacteria in these classification models also supports a role for bacteria in colorectal cancer. Here we built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared performance to models built using bacterial community signatures.

We built and tested random forest models to classify stool samples as belonging to either cancerous or healthy individuals. These models were based on virus metagenomic community signatures or bacterial 16S rRNA gene signatures. We also included a whole metagenomic sequence set to ensure viral metagenomic observations were not a trait of metagenomics in general. Each operational units' relative abundance was used in the feature set. The same model approach was used for all three datasets, and the only difference was the data used to learn the model. We confirmed that our model using bacterial 16S data replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba].

We compared the bacterial 16S rRNA gene model to a model built using the virome signatures. The viral model performed as well as the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively, after fifteen random forest iterations **(Figure \ref{predmodel} A - B)**.

To confirm that this observation was due to the virome community itself, and was not a trait of metagenomic datasets in general, we built a model using whole metagenomic community signatures. This model performed poorly with a mean AUC of `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)`, lending support to our observation that classification performance is a trait unique to the virus metagenome **(Figure \ref{predmodel} A - B)**. Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and lowly abundant OTUs. Further filtration of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**.

To evaluate the synergistic capabilities of the bacterial and viral signatures within the model, we built a combinatory model using both bacteria community and virome data. The combination model yielded modest but significantly improved performance beyond the virome (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Combined"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, yielding an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**. This suggests the virus and bacterial communities may have synergistic capabilities for classifying stool as belonging to cancerous individuals.

The association between the two communities and colorectal cancer was driven by a few important microbes, measured using the mean decrease in model accuracy when each was iteratively removed. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with it's previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature was also driven primarily by a few operational genomic units, suggesting a role in cancer development **(Figure \ref{predmodel} D)**. The identified important viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and orphan phage taxa without taxonomic identifiers (denoted "unclassified"). Many of the important viruses were unidentifiable (denoted "unknown"), suggesting they are members of the abundant viral dark matter (uncharacterized virus genomic space) associated with the human virome. This is common in the virome, and studies can have as much as 95% of virus sequences belong to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz].

When the bacterial and viral community signatures were combined, two bacterial and two viral organisms primarily drove the community association with cancer **(Figure \ref{predmodel} E)**. The most important microbes were two unidentified viruses, followed by *Bacteroides* and *Fusobacterium*.

## The colon Virome Allows for Cancerous, Adenomatous, and Healthy Classification
After evaluating our ability to classify samples as cancerous or healthy, we incorporated the precancerous adenoma samples into the model and evaluated our ability to classify all three states in our total sample set. We used three-class random forest models for the bacterial 16S and viral sample sets. The bacterial signature model yielded an AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded an overall AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**). Both models were best able to classify cancer samples from healthy or precancerous samples, but struggled to distinguish precancerous from healthy or cancerous **(Figure \ref{threewaymodel} A-B)**. The cancerous signal was the most discriminatory of the three sample types.

Many of the microbes important for the two-class (cancer vs healthy) bacteria and virus models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**. The most important bacterium was the same *Fusobacterium* between the two and three class models, supporting its significance to the association between cancer and the bacterial communities **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**. Unlike the two-class virome model, the viruses most important to the three-class model were identified bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**. A *Podoviridae* was most important, followed by two *Siphoviridae* phages.

@@

## Temperate Bacteriophages Drive Link Between Virome and Colorectal Cancer
The virome-based model was able to accurately classify stool samples as cancerous, precancerous, or healthy. Not only is this important for establishing an alternative diagnostic model, but it also suggests an underlying biological importance for viruses in colorectal cancer. Above we used our classification models to evaluate which virus OGUs were most highly linked to colorectal cancer. We were able to further characterize these important OGUs to better understand the underlying biological link between the virome and colorectal cancer.

The role of the virome in colorectal cancer could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts. To better understand the types of viruses that are important for colorectal cancer, we utilized the longest sequences from the OGUs as the representative sequences to be used for taxonomic classification. These sequences were aligned to a set of all reference virus genomes, including bacteriophages and eukaryotic viruses. A strict e-value threshold of 1e-25 was used to improve our confidence in the matches between the genome sequences. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel} E)**. Overall we were able to identify `r signif(idcount, digits = sigfig)`% of the OGUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OGUs aligned to bacteriophage reference genomes.

We evaluated whether the phages in the community were primarily lytic (lyse their hosts) or temperate (lysogenic; can integrate into their host's genome, as well as lyse the cell). We accomplished this by identifying three markers for temperate phages in the OGU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes. This approach was done as described in previous work [@Minot:2011ez; @Hannigan:2015fz]. We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{threewaymodel} E)**. Thus the majority of the OGUs are temperate bacteriophages and not eukaryotic viruses, indicating the association between the virome and colorectal cancer is reliant on bacteriophage communities that can lie dormant in bacterial genomes. These findings are consistent with previous reports suggesting the colon virome is primarily temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].

## Influential Bacteria and Phages Associate Indirectly
Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the influential phages were predators of the influential bacteria, and thus influenced their relative abundance through predation (e.g. One of the important phages targets *Fusobacterium*). If this hypothesis was true, we would expect that the relative abundance between many important bacteria and phages to be correlated. We correlated the relative abundance values of every phage OGU and bacterial OTU and found strikingly low correlation between the bacterial and phage abundances **(Figure \ref{correlations} A,C)**. There was an overall absence of correlations between the most influential phage OGUs and bacterial OTUs **(Figure \ref{correlations} B)**. This evidence supported our null hypothesis that the influential phages are not predators of the influential bacterial.

>>>>

Colorectal cancer is the second leading cause of cancer-related death in the United States and is a primary cause of morbidity and mortality throughout the world. Although the majority of colorectal cancer case causes remains unclear, it's progression has been linked to colonic bacterial community composition. Viruses are another important component of the colonic microbial community that have yet to be studied in colorectal cancer, despite their oncogenic potential. We evaluated the colorectal cancer virome (virus community) using a cohort of 90 human subjects with either healthy, adenomatous (precancerous), or cancerous colons. We utilized 16S rRNA gene, whole shotgun metagenomic, and purified virus metagenomic sequencing methods to compare the colorectal cancer virome to the bacterial community. We found that alpha and beta diversity metrics were insufficient for detecting virome changes in colorectal cancer, but more sophisticated virome-based random forest models identified significant changes in the virus community. The majority of the cancer-associated virome consisted of temperate bacteriophages, suggesting the community was indirectly linked to colorectal cancer by modulating bacterial community structure and functionality. These results provide foundational evidence that bacteriophage communities are associated with colorectal cancer and likely impact cancer progression by altering the bacterial host communities. Together our findings add to the existing model for the role the microbiome plays in colorectal cancer development, thus providing us with a more complete understanding of colorectal cancer etiology.

**Word Count**: 226

# Significance Statement
Colorectal cancer is a leading cause of cancer-related death in the United States and worldwide. Its progression and severity have been linked to colonic bacterial community composition. Little is known about cancer-associated colon virus communities and their influence on bacteria. We began addressing this knowledge gap by identifying changes in colonic virus communities in colorectal cancer patients, and how they compare to bacterial community changes. The altered colorectal cancer virome was used in virome-based machine learning models to accurately classify patient cancer status. The cancer-virus classifier was driven primarily by bacteriophages (bacterial viruses). Wider phage host ranges were more strongly associated with colorectal cancer. The results suggest an indirect role for the virome impacting colorectal cancer by modulating their associated bacterial community.

**Word Count**: 125

@@

# Introduction
Due to their mutagenic abilities and their propensity for functional manipulation, human viruses are strongly associated with, and in many cases cause, cancer [@Feng:2008kr; @Shuda:2011gf; @Schiller:2012ba; @Chang:1994up]. Because bacteriophages are crucial for bacterial community stability and composition [@Harcombe:2005fd; @RodriguezValera:2009cr; @Cortez:2014bk], and because bacteria have been implicated as oncogenic agents [@Zackular:2014fba; @Garrett:2015fg; @Baxter:2014hb], bacteriophages have the potential to indirectly impact cancer. The gut virome (the virus community of the gut) therefore has the potential to impact health and disease (e.g. cancer), and has been associated with diseases including periodontal disease [@Ly:2014ew], HIV [@Monaco:2016ita], antibiotic exposure [@Abeles:2015dy; @Modi:2013fi], urinary tract infections [@SantiagoRodriguez:2015gd], and inflammatory bowel disease [@Norman:2015kb]. The precedence of the virome impacting human health and the strong association of bacterial communities with colorectal cancer suggest that colorectal cancer may be linked to altered virus communities.

Colorectal cancer is the second leading cause of cancer-related deaths in the United States [@Siegel:2014jo]. The US National Cancer Institute estimates over 1.5 million Americans have been diagnosed with colorectal cancer in 2016, and over 500,000 Americans will have died from the disease [@Siegel:2014jo]. Although the majority of colorectal cancer case causes remains unclear, variation in colorectal bacterial communities have been linked to the disease [@Flynn:2016iu; @Zackular:2016en; @Zackular:2014fba; @Baxter:2014hb]. This work has led to the proposal of disease models in which bacteria colonize the colon, develop biofilms, promote infiltration, and enter an oncogenic synergy with the cancerous human cells [@Flynn:2016iu]. This association has also allowed researchers to leverage bacterial community signatures as biomarkers to provide accurate, noninvasive colorectal cancer detection from stool [@Zackular:2014fba; @Baxter:2016dja]. While an understanding of colorectal cancer bacterial communities has proven fruitful both for disease classification and understanding underlying etiology, bacteria are only a subset of the colon microbiome. Viruses are another important component of the colon microbial community that have yet to be studied in the context of colorectal cancer. We evaluated colorectal cancer disruptions in virus and bacterial community composition in a human cohort whose stool was sampled at the three relevant stages of cancer developement: healthy, adenomatous, and cancerous.

Colorectal cancer is a stepwise process that begins when healthy tissue develops into a precancerous polyp (i.e. adenoma) in the large intestine [@Fearon:2011go]. If left untreated, the adenoma will develop into a cancerous lesion that can invade and metastasize, leading to severe illness and death. Progression to cancer can be prevented when precancerous adenomas are detected and removed during routine screening [@Levin:2008kz; @Zauber:2015cr], and survival for colorectal cancer patients may exceed 90% when the lesions are detected early and removed [@Levin:2008kz]. Thus work such as this, which aims to power early detection and prevention of progression from early cancer stages, has a great potential to inform therapeutic development.

In this study we report that colorectal cancer is associated with variation in colonic virus communities. The majority of viruses identified within the virome were temperate bacteriophages. Just as the association between the bacterial community and colorectal cancer was driven by select influential bacteria including *Fusobacterium*, the association between the virome and colorectal cancer was also driven by a subset of influential phages. Our data suggest that the influential phages do not exclusively infect influential bacterial (e.g. *Fusobacterium* phages infecting the *Fusobacterium*), but rather act through the community as a whole. The implications of these findings are threefold. *First*, this supports a biological role for the virome in colorectal cancer development and suggests that more than bacteria are involved in the process. *Second*, we present a supplementary, or even alternative virus-based approach for classification modeling of colorectal cancer using stool samples. *Third*, we provide initial support for the importance of studying the virome as a component of the microbiome ecological network, especially in cancer.

@@

## The Colorectal Cancer Virome Cohort
The study cohort consisted of 90 human subjects, 30 of which had healthy colons, 30 of which had adenomas, and 30 which had carcinomas **(Figure \ref{sampleproc})**. Half of the stool was used to sequence the bacterial communities using both 16S rRNA gene and shotgun sequencing techniques. The other half of the stool samples were purified for virus like particles (VLPs) and consequent genomic DNA extraction, followed by shotgun metagenomic sequencing. The VLP purification allowed us to observe the *active virome* because we only sequenced those viruses that are encapsulated.

Virus DNA was purified before sequencing, which allowed us to analyze DNA from within virus capsids **(Figure \ref{sampleproc})**. Each extraction protocol was performed with a blank control to detect any contaminants from reagents. Only one of the nine controls contained detectable DNA, which was of a minimal concentration, thus providing initial evidence of successful sequencing of VLP genomic DNA over potential contaminants **(Figure \ref{qualcontrol} A)**. As was expected, these controls were sparsely sequenced and were mostly removed while sub-sampling to even depths **(Figure \ref{qualcontrol} B)**. The high quality phage and bacterial sequences were assembled into highly covered contigs longer than 1kb **(Figure \ref{contigqc})**. Because contigs only represent genome fragments, we further clustered related contigs into operational genomic units (OGUs; conceptually similar to 16S rRNA gene operational taxonomic units) with the majority containing hundreds of related contigs **(Figure \ref{contigqc} - \ref{clustercontigqc})**.

## Virus Diversity and Colorectal Cancer
Microbiome and disease associations are often described as being of an altered diversity (i.e. "dysbiotic"). We therefore initially evaluated the diversity of virome OGUs and their association with colorectal cancer. We utilized the Bray-Curtis dissimilarity metric to evaluate differences in communities between disease states. To control for uneven sequencing depths, we subsampled to a minimum depth that maintained most samples while excluding the sparsely sequenced blank controls.

We calculated the differences in virus alpha diversity associated with colorectal cancer. We found no significant alterations in either Shannon entropy or richness **(Figure \ref{betaogu} C-D)**. There was also no statistically significant clustering of the disease groups (ANOSIM p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithoutneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogu}**). It is worth noting that there was a significant difference between the blank controls (those few that remained after sub-sampling) and the other study groups, further supporting the quality of our sample set (Anosim p-value = `r signif(alphadf[c(alphadf$Title %in% "anovawithneg"),"Stat"], digits = sigfig)`, **Figure \ref{betaogunegative})**. Overall, standard diversity metrics were insufficient for capturing the differences in the virus communities between disease states.

## Virome Composition Changes in Colorectal Cancer
Previous work has shown that 16S rRNA gene relative abundance profiles are effective features for classifying stool samples as originating from healthy, adenomatous, or cancerous individuals [@Zackular:2014fba; @Baxter:2016dja]. The exceptional performance of bacteria in these classification models supports a role for bacteria in colorectal cancer. Here we built off of these findings by evaluating the ability of virus community signatures to classify stool samples and compared performance to models built using bacterial community signatures.

We built and tested random forest models to classify stool samples as belonging to either cancerous or healthy individuals. These models were based on virus metagenomic community or bacterial 16S rRNA gene relative abundance profiles. We confirmed that our bacterial 16S rRNA gene model replicated the performance of the original report which used logit models instead of random forest models **(Figure \ref{predmodel} A)** [@Zackular:2014fba]. We then compared the bacterial 16S rRNA gene model to a model built using virome relative abundance. The viral model performed as well as the bacterial model (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) with the viral and bacterial models achieving mean AUC (area under the curve) values of `r signif(twoauc[c(twoauc$class %in% "Virus"),"meanAUC"], digits = sigfig)` and `r signif(twoauc[c(twoauc$class %in% "Bacteria"),"meanAUC"], digits = sigfig)`, respectively **(Figure \ref{predmodel} A - B)**.

To evaluate the synergistic capabilities of the bacteria and viruses within the model, we built a combinatory model that used both bacterial and viral community data. The combination model yielded modest but significantly improved performance beyond the separate virome (corrected p-value = `r signif(twoss[c(twoss$First %in% "Virus" & twoss$Second %in% "Combined"),"pval"], digits = sigfig)`) and bacterial (corrected p-value = `r signif(twoss[c(twoss$First %in% "Combined" & twoss$Second %in% "Bacteria"),"pval"], digits = sigfig)`) models, and yielded an AUC of `r signif(twoauc[c(twoauc$class %in% "Combined"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**. This suggested that features from the virus and bacterial communities may have had synergistic capabilities for classifying stool as belonging to cancerous individuals.

We also compared these models to relative abundance profiles from bacterial metagenomic shotgun sequencing data. This model performed poorly, with a mean AUC of `r signif(twoauc[c(twoauc$class %in% "Metagenomic"),"meanAUC"], digits = sigfig)` **(Figure \ref{predmodel} A - B)**. Further investigation revealed that the bacterial 16S rRNA gene model was strongly driven by sparse and lowly abundant OTUs **(Figure \ref{16scompare})**. Filtration of OTUs with a median abundance of zero resulted in the removal of six OTUs, and a loss of model performance down to what was observed in the metagenome **(Figure \ref{16scompare} A)**. The majority of these OTUs had a relative abundance lower than 1% **(Figure \ref{16scompare} B)**.

The association between the two communities and colorectal cancer was driven by a few important microbes, measured using the mean decrease in model accuracy when each was removed. *Fusobacterium* was the primary driver of the bacterial association with colorectal cancer, which is consistent with its previously described oncogenic potential **(Figure \ref{predmodel} C)**[@Flynn:2016iu]. The virome signature was also driven by a few operational genomic units, suggesting a role for the OGUs in cancer development **(Figure \ref{predmodel} D)**. The identified viruses were bacteriophages, belonging to *Siphoviridae*, *Myoviridae*, and orphan phage taxa without taxonomic identifiers (denoted "unclassified"). Many of the important viruses were unidentifiable (denoted "unknown"), suggesting they are members of the abundant unknown viral population associated with the human virome. This is common in the virome; studies can have as much as 95% of virus sequences belong to unknown genomic units [@Pedulla:2003tu; @Hannigan:2015fz]. When the bacterial and viral community signatures were combined, both bacterial and viral organisms drove the community association with cancer **(Figure \ref{predmodel} E)**.

## Phage Influence Shifts During CRC Progression
Because our cohort included healthy, adenomatous, and cancerous colons, we were able to gain insight into virus community shifts during cancer progression. We evaluated community shifts between the two disease transitions (healthy to adenomatous and adenomatous to cancerous) by building random forest models to compare only the sample classes around the transitions. We found that, while bacterial 16S rRNA gene models perform equally well for all disease class comparisons, the virome models performed worse when seperating healthy from adenomatous, and better when seperating adenomatous from cancerous **(Figure \ref{transitionmodels} A-B)**. Like bacteria **(Figure \ref{transitionmodels} F-H)**, different virome members drove the transitions from healthy to adenomatous and adenomatous to cancerous, with one phage exception **(Figure \ref{transitionmodels} C-E)**. A *Myoviridae* OGU (Cluster 188) was the seventh most important phage driving the transition from a healthy to adenomatous state, and was the most important phage driving the progression from an adenomatous to carcingenic state. Therefore, like bacteria, there are distinct phages assocaited with the transitions from a healthy to an adenomatous colon, and from an adenomatous to a cancerous colon.

## Inclusive Colonic Virome Classification
After evaluating our ability to classify samples between two disease states, we performed a comprehensive three-class classification of all three disease states. We used a three-class random forest model for the bacterial 16S rRNA gene and viral sample sets. The bacterial signature model yielded an AUC of `r signif(threeauc[c(threeauc$class %in% "Bacteria"),"meanauc"], digits = sigfig)` and outperformed the viral community model which yielded an overall AUC of `r signif(threeauc[c(threeauc$class %in% "Virus"),"meanauc"], digits = sigfig)` (p-value = `r signif(threesig, digits = sigfig)`, **Figure \ref{threewaymodel} A-C**). Both models were best able to classify cancer samples from healthy or precancerous samples, but struggled to distinguish adenomatous from healthy or cancerous **(Figure \ref{threewaymodel} A-B)**. The cancerous signal was the most discriminatory of the three sample types.

The microbes important for the cancer vs healthy and healthy vs adenoma bacteria and virus models were also important for the three-class model **(Figure \ref{threewaymodel} D-E)**. The most important bacterium was the same *Fusobacterium* between the two and three class models, supporting its significance to the association between cancer and the bacterial communities **(Figure \ref{predmodel} C, Figure \ref{threewaymodel} D)**. The viruses most important to the three-class model were identified as bacteriophages **(Figure \ref{predmodel} D, Figure \ref{threewaymodel} E)**.

@@

## Dominance of Bacteriophages in Colorectal Cancer Virome
Changes in the colorectal cancer virome could have been driven directly by eukaryotic viruses or indirectly by bacteriophages acting through their bacterial hosts. To better understand the types of viruses that are important for colorectal cancer, we identified the virome OGUs as being similar to either eukaryotic viruses or bacteriophages. The most important viruses to the classification model were identified as bacteriophages (**Figure \ref{threewaymodel})**. Overall we were able to identify `r signif(idcount, digits = sigfig)`% of the OGUs as known viruses, and `r signif(phagecount, digits = sigfig)`% of those viral OGUs aligned to bacteriophage reference genomes.

We evaluated whether the phages in the community were primarily lytic (replicate by lysing their host) or temperate (lysogenic; able to integrate into their host's genome, as well as lyse the cell). We accomplished this by identifying three markers for temperate phages in the OGU representative sequences: 1) presence of phage integrase genes, 2) presence of known prophage genes, according the the ACLAME (A CLAssification of Mobile genetic Elements) database, and 3) nucleotide similarity to regions of bacterial genomes. This approach was done as described in previous work [@Minot:2011ez; @Hannigan:2015fz]. We found that the majority of the colon phages were temperate, and that the overall fraction of temperate phages remained consistent throughout the healthy, adenomatous, and cancerous stages **(Figure \ref{replicationstyles} E)**. Thus the majority of the OGUs are temperate bacteriophages and not eukaryotic viruses, indicating the association between the virome and colorectal cancer is reliant on bacteriophage communities that can lie dormant in bacterial genomes. These findings are consistent with previous reports suggesting the gut virome is primarily temperate phages [@Minot:2011ez; @Monaco:2016ita; @Norman:2015kb; @Reyes:2010cwa].

## Community Context of Influential Phages
Because the link between colorectal cancer and the virome was driven by bacteriophages, we hypothesized that the influential phages were predators of the influential bacteria, and thus influenced their relative abundance through predation. If this hypothesis were true, we would expect the relative abundance between important bacteria and phages to be correlated. Instead we observed a strikingly low correlation between the bacterial and phage relative abundances **(Figure \ref{correlations} A,C)**. There was an overall absence of correlations between the most influential phage OGUs and bacterial OTUs **(Figure \ref{correlations} B)**. This evidence supported our null hypothesis that the influential phages are not primarily predators of influential bacteria.

Our next hypothesis was that the influential phages were acting by infecting a wide range of bacteria in the overall community, instead of just the select influential bacteria. This hypothesis would be supported by showing that the influential bacteriophages were community hubs within the bacteria and phage interactive network. We investigated the infectious capabilities and potential host ranges of the influential phages by building a machine learning model to predict which phage OGUs infect which bacteria in the overall community. The predicted interactions were then used to build a network which was used for subsequent analysis, as has been previously described (**Cite Network Preprint**). This analysis revealed a wide tropism range for the bacteriophages within the community **(Figure \ref{network} A)**. We calculated the alpha centrality (measure of importance in the ecosystem network) of each phage OGU's connection to the rest of the network, and compared the centrality to the importance of each OGU in the colorectal cancer classification model. We found that phage OGU centrality is significantly positively correlated with importance to the disease model (p-value = `r signif(corstats[c(corstats$names %in% "pval"),"scores"], digits = sigfig)`, rho = `r signif(corstats[c(corstats$names %in% "rho"),"scores"], digits = sigfig)`), indicating that phages important in driving colorectal cancer were also community hubs **(Figure \ref{network} B)**. Together these findings supported our hypothesis that influential phages were hubs within their microbial communities.
